Scalable Clustering of Genotype Information using MapReduce

### 2015-03-11 doc/publications/journal2014/paper.tex
Corrected figure numbers

\subsection*{Result visualisation}
% DB: in the conclusion we discussed that we provide a "toolset to interpret Mahout's clustering output". 
% DB: can we discuss here how we get a visual representation from the clustering (ClusterDumper+Gephi) and what the image tells us..
GraphML files generated using Mahout Cluster Dumper can be visualised using popular freely-available tools such as Gephi~\cite{ICWSM09154}. An example visualisation using the phase 3 dataset, with nodes coloured by super population, is shown in Figure XX.

@@

\section*{Conclusion}
In this paper we have applied Mahout's k-means clustering to the task of clustering individuals, based on their genomic variants. 
We provide an interface from VCF format to Hadoop and Mahout, and a toolset to interpret Mahout's clustering output (see Figure 1)
We find that on small datasets the clustering stage is two orders of magnitudes slower in Mahout compared to R (9 minutes vs 7 seconds), due to the overhead of setting up and co-ordinating the communication between the Hadoop nodes. 
However, increasing either the number of samples or the number of features per sample quickly exceeds the capability of R on this machine due to memory limitations.
%DB ok effortlessly might be overstated ?
Hadoop on the other hand scales effortlessly to more than twice the number samples (2,504). It can also process VCF files that contain over 79 Million variants per samples, albeit with higher memory allocation.
The MapReduce functions can scale from a relatively small and accessible cluster with 16 cores, to a much larger cluster with close to 500 CPU cores. 
In conclusion, Mahout offers a scalable out-of-the box solution that is directly applicable to genotype-based clustering for population or family association.

>>>>

\subsection*{Result visualisation}
% DB: in the conclusion we discussed that we provide a "toolset to interpret Mahout's clustering output". 
% DB: can we discuss here how we get a visual representation from the clustering (ClusterDumper+Gephi) and what the image tells us..
GraphML files generated using Mahout Cluster Dumper can be visualised using popular freely-available tools such as Gephi~\cite{ICWSM09154}. An example visualisation using the phase 3 dataset, with nodes coloured by super population, is shown in Figure 1.

@@

\section*{Conclusion}
In this paper we have applied Mahout's k-means clustering to the task of clustering individuals, based on their genomic variants. 
We provide an interface from VCF format to Hadoop and Mahout, and a toolset to interpret Mahout's clustering output (see Figure 2)
We find that on small datasets the clustering stage is two orders of magnitudes slower in Mahout compared to R (9 minutes vs 7 seconds), due to the overhead of setting up and co-ordinating the communication between the Hadoop nodes. 
However, increasing either the number of samples or the number of features per sample quickly exceeds the capability of R on this machine due to memory limitations.
%DB ok effortlessly might be overstated ?
Hadoop on the other hand scales effortlessly to more than twice the number samples (2,504). It can also process VCF files that contain over 79 Million variants per samples, albeit with higher memory allocation.
The MapReduce functions can scale from a relatively small and accessible cluster with 16 cores, to a much larger cluster with close to 500 CPU cores. 
In conclusion, Mahout offers a scalable out-of-the box solution that is directly applicable to genotype-based clustering for population or family association.
